Prognostic and biological relevance of collagen-related genes in prostate cancer

### 2024-01-25 report/markdown/manuscript_figures_tables.Rmd
ISUP grading in Figure 1

Differential gene expression in cancer samples stratified by Gleason scoring (5 - 6, 7 and $\geq$ 8) was assessed by FDR-corrected one-way ANOVA with $\eta^2$ effect size statistic. 
Genes with pFDR < 0.05 and $\eta^2 \geq$ 0.02 were deemed differentially regulated (__Supplementary Table S4__).

@@

_(C) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` collagen-related genes between cancer samples stratified by Gleason scores were assessed by one-way ANOVA with $\eta^2$ effect size statistic in the GSE54460 (`r format_strata_n(gs_uni$n_numbers$gse54460)`), GSE70768 (`r format_strata_n(gs_uni$n_numbers$gse70768)`), GSE70769 (`r format_strata_n(gs_uni$n_numbers$gse70769)`), GSE220095 (`r format_strata_n(gs_uni$n_numbers$gse220095)`), DKFZ (`r format_strata_n(gs_uni$n_numbers$dkfz)`), and TCGA cohort (`r format_strata_n(gs_uni$n_numbers$tcga)`). Mean normalized $log_2$-transformed expression levels of genes significantly regulated between the Gleason score strata with at least weak effect size (pFDR < 0.05, $\eta^2 \geq$ 0.02) are presented in a heat map._

>>>>

Differential gene expression in cancer samples stratified by the ISUP risk strata (ISUP1: Gleason score 5 - 6, ISUP2: Gleason score 7 and ISUP3: Gleason score $\geq$ 8) was assessed by FDR-corrected one-way ANOVA with $\eta^2$ effect size statistic. 
Genes with pFDR < 0.05 and $\eta^2 \geq$ 0.02 were deemed differentially regulated (__Supplementary Table S4__).

@@

_(C) Differences in $log_2$-transformed expression levels of `r nrow(globals$genes_interest)` collagen-related genes between cancer samples stratified according to the ISUP risk system were assessed by one-way ANOVA with $\eta^2$ effect size statistic in the GSE54460 (`r format_strata_n(gs_uni$n_numbers$gse54460)`), GSE70768 (`r format_strata_n(gs_uni$n_numbers$gse70768)`), GSE70769 (`r format_strata_n(gs_uni$n_numbers$gse70769)`), GSE220095 (`r format_strata_n(gs_uni$n_numbers$gse220095)`), DKFZ (`r format_strata_n(gs_uni$n_numbers$dkfz)`), and TCGA cohort (`r format_strata_n(gs_uni$n_numbers$tcga)`). Mean normalized $log_2$-transformed expression levels of genes significantly regulated between the Gleason score strata with at least weak effect size (pFDR < 0.05, $\eta^2 \geq$ 0.02) are presented in a heat map._

### 2024-01-25 report/markdown/manuscript_supplement.Rmd
ISUP grading in Figure 1

## Comparison of gene expression between the benign and cancer tissue, and cancer samples stratified by Gleason scores

Differences in $log_2$-transformed expression levels of the collagen pathway genes between donor-matched cancer and benign prostate specimens were investigated with paired two-tailed T test with Cohen's effect size statistic in the GSE70768 and TCGA cohorts (__Figure 1B__, __Supplementary Figure S\@ref(fig:fig-norm-tumor)__ and __Supplementary Table S\@ref(tab:tab-norm-tumor)__). 
Differences in gene expression between cancer samples stratified by Gleason score (5 - 6, 7, $\geq$ 8) were explored by one-way ANOVA with $\eta^2$ effect size statistic (__Figure 1C__, __Supplementary Table S\@ref(tab:tab-gleason)__).
Differentially regulated genes were defined by pFDR < 0.05 and at least weak effect size of the difference (d $\geq$ 0.2 or $\eta^2 \geq$ 0.02). 
The analyses were done with the function `compare_variables()` from the [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package.

@@

Biochemical relapse-free survival in the pooled GEO training cohort was modeled with normalized (Z-scores) expression values of the the collagen-related genes as explanatory variables by the GBM algorithm [@Friedman2001; @Greenwell2022; @Natekin2013; @Friedman2002]. 
Selection of the optimal set of modeling parameters (number of trees, shrinkage factor, interaction depth, and minimal number of observations in the tree node) was accomplished by tuning with 10-fold cross-validation and out-of-fold model deviance as performance statistic (__Supplementary Table S\@ref(tab:tab-tuning)__). 
The tuning was performed with an in-house developed script. 
The GBM model was trained in the pooled GEO cohort with the optimal tuning parameter set with the `gbm()` function (package _gbm_). 
The Transcriptomic Collagen Score was defined as the linear predictor score of the GBM model and was computed for the best model iteration (function `gbm.perf(method = 'cv')`) with the `predict()` method and expression data from the training pooled GEO cohort and the TCGA and DKFZ test collectives. 
Importance of the explanatory variables for the predictive performance of the GBM model in the training pooled GEO cohort were measured with the relative influence method [@Friedman2001] and expressed as gradient of the sum of squared errors attributed to particular variables ($\Delta SSE$, __Figure 3A__).

@@

```{r tab-gleason, tab.cap = 'Expression of the collagen pathway genes in cancer samples stratified by Gleason scores compared by one-way ANOVA with eta-square effect size statistic. P values were corrected for multiple testing with the false discovery rate method. log2-transformed expression values are presented as medians with interquartile ranges (IQR) and ranges. The table is available as a supplementary Excel file.'}

>>>>

## Comparison of gene expression between the benign and cancer tissue, and cancer samples stratified by the ISUP risk scheme

Differences in $log_2$-transformed expression levels of the collagen pathway genes between donor-matched cancer and benign prostate specimens were investigated with paired two-tailed T test with Cohen's effect size statistic in the GSE70768 and TCGA cohorts (__Figure 1B__, __Supplementary Figure S\@ref(fig:fig-norm-tumor)__ and __Supplementary Table S\@ref(tab:tab-norm-tumor)__). 
Differences in gene expression between cancer samples stratified according to the ISUP risk scheme (ISUP1: Gleason Score 5 - 6, ISUP2: Gleason score 7, ISUP3+: Gleason Score $\geq$ 8) were explored by one-way ANOVA with $\eta^2$ effect size statistic (__Figure 1C__, __Supplementary Table S\@ref(tab:tab-gleason)__).
Differentially regulated genes were defined by pFDR < 0.05 and at least weak effect size of the difference (d $\geq$ 0.2 or $\eta^2 \geq$ 0.02). 
The analyses were done with the function `compare_variables()` from the [_ExDA_](https://github.com/PiotrTymoszuk/ExDA) package.

@@

Biochemical relapse-free survival in the pooled GEO training cohort was modeled with normalized (Z-scores) expression values of the collagen-related genes as explanatory variables by the GBM algorithm [@Friedman2001; @Greenwell2022; @Natekin2013; @Friedman2002]. 
Selection of the optimal set of modeling parameters (number of trees, shrinkage factor, interaction depth, and minimal number of observations in the tree node) was accomplished by tuning with 10-fold cross-validation and out-of-fold model deviance as performance statistic (__Supplementary Table S\@ref(tab:tab-tuning)__). 
The tuning was performed with an in-house developed script. 
The GBM model was trained in the pooled GEO cohort with the optimal tuning parameter set with the `gbm()` function (package _gbm_). 
The Transcriptomic Collagen Score was defined as the linear predictor score of the GBM model and was computed for the best model iteration (function `gbm.perf(method = 'cv')`) with the `predict()` method and expression data from the training pooled GEO cohort and the TCGA and DKFZ test collectives. 
Importance of the explanatory variables for the predictive performance of the GBM model in the training pooled GEO cohort were measured with the relative influence method [@Friedman2001] and expressed as gradient of the sum of squared errors attributed to particular variables ($\Delta SSE$, __Figure 3A__).

@@

```{r tab-gleason, tab.cap = 'Expression of the collagen pathway genes in cancer samples stratified by the ISUP risk groups compared by one-way ANOVA with eta-square effect size statistic. P values were corrected for multiple testing with the false discovery rate method. log2-transformed expression values are presented as medians with interquartile ranges (IQR) and ranges. The table is available as a supplementary Excel file.'}
